---
title: "Final Project DS 202"
author: "Minh Nguyen, Bach Nguyen, Thien Nguyen"
date: "2024-12-17"
output: github_document  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract

This study analyzes mobile device usage patterns and user behavior classifications based on various metrics such as **App Usage Time**, **Screen-On Time**, **Battery Drain**, **Data Usage**, and **Number of Apps Installed**. The dataset includes 700 users categorized into **five behavior classes**, ranging from **light (Class 1)** to **extreme usage (Class 5)**. Results show that higher behavior classes are associated with increased Screen-On Time, Battery Drain, and Data Usage, highlighting strong positive relationships among these usage metrics. Gender differences were observed, with **males** generally dominating higher behavior classes, while females in **Class 5** displayed slightly higher Screen-On Time. Age, however, showed no significant correlation with usage metrics. These findings provide insights into user behavior trends, emphasizing how device usage patterns escalate with user behavior intensity.

## Main Analysis - Curiosity

The analysis demonstrates a deep exploration of mobile device usage patterns, focusing on key metrics like Screen-On Time, App Usage Time, and Battery Drain to understand user behavior classes. Several trials were conducted to identify trends, including examining correlations, gender differences, and age distributions. The findings were refined to present the most meaningful insights, such as the strong positive relationships between usage metrics and behavior intensity. Additional research and visualizations were used to validate these observations, providing a clear explanation of how user behavior escalates across different classes and highlighting areas for further investigation.

## Main analysis - Skepticism

The analysis critically examined the findings by using multiple approaches, such as correlation heatmaps, boxplots, and histograms, to verify relationships between variables. Unintuitive results, like the lack of correlation between **Age** and other usage metrics, were cross-checked with different visualizations to ensure accuracy. The consistency of trends, such as the strong positive correlations between **Screen-On Time**, **Battery Drain**, and **Data Usage**, was validated across various techniques. This careful approach ensured the results were reliable and reflected true user behavior patterns.

## Main analysis - Organization

The analysis of mobile device usage patterns focuses on exploring the relationships between **User Behavior Classes** and key metrics, such as **Screen-On Time**, **Battery Drain**, **Data Usage**, and **Number of Apps Installed**, while also considering **Gender** and **Age Groups**. Users were categorized into five classes, ranging from **light usage (Class 1)** to **extreme usage (Class 5)**. Boxplots revealed that **Screen-On Time**, **Battery Drain**, and **Data Usage** increased steadily with higher behavior classes, suggesting that these metrics are strong indicators of usage intensity. Gender-based analysis showed that males generally dominate heavier behavior classes, but in **Class 5**, females exhibited slightly higher **Screen-On Time**.  

Additionally, the data was analyzed across **Age Groups** to understand variations within behavior classes. Results showed that **Screen-On Time** increased consistently with behavior classes across all age groups, with no major differences among age categories. Histograms of key metrics like App Usage Time, Battery Drain, and Data Usage revealed right-skewed distributions, where most users had lower values, but a few exhibited very high usage. The correlation heatmap confirmed strong positive relationships between metrics like **Screen-On Time**, **App Usage Time**, and **Battery Drain**, while **Age** showed no significant correlation with any metric. This systematic analysis highlights the strong connections between user behavior classes and device usage metrics, providing a clear understanding of patterns and trends across different user groups.

## Introduction

Understanding mobile device usage patterns is crucial in today's digital age, where smartphones play a significant role in daily life. This study explores user behavior by analyzing metrics such as **App Usage Time**, **Screen-On Time**, **Battery Drain**, **Number of Apps Installed**, and **Data Usage**. Users are categorized into **five behavior classes**, ranging from **light usage (Class 1)** to **extreme usage (Class 5)**. The analysis highlights key trends, such as increased Screen-On Time and Data Usage with higher behavior classes, as well as slight gender differences in usage intensity. By uncovering these patterns, this study provides insights into user habits and lays the groundwork for developing predictive models to classify behavior, optimize app performance, and improve digital well-being strategies.

## Importing libraries

```{r}
library(RColorBrewer)
library(viridis)
library(psych)
library(dplyr)
library(ggplot2)
library(patchwork)
library(tidyr)
library(corrplot)
library(png)
library(magick)
library(ggcorrplot)
library(gridExtra)
```

## Data

The dataset contains **700 records** of mobile device usage behavior, with users categorized into **five behavior classes** ranging from **light usage (Class 1)** to **extreme usage (Class 5)**. Each record includes the following features:  

1. **App Usage Time (minutes/day)**: Total time spent using mobile applications each day.  
2. **Screen-On Time (hours/day)**: The number of hours the device screen remains active per day.  
3. **Battery Drain (mAh/day)**: The daily battery consumption in milliampere-hours.  
4. **Number of Apps Installed**: Total number of applications installed on the device.  
5. **Data Usage (MB/day)**: The total amount of mobile data consumed per day.  
6. **Age**: Age of the user in years.  
7. **Gender**: Gender of the user, categorized as **Male** or **Female**.  
8. **User Behavior Class**: A categorical variable representing user behavior intensity, with values:  
   - **1**: Light usage  
   - **2**: Light-to-moderate usage  
   - **3**: Moderate usage  
   - **4**: Heavy usage  
   - **5**: Extreme usage  

This dataset allows for the analysis of relationships between user behavior classes and various device usage metrics, enabling the identification of trends and patterns across different user groups.


Data link: <https://www.kaggle.com/datasets/valakhorasani/mobile-device-usage-and-user-behavior-dataset/data>

## Load & Exploring Data

```{r}
df <- read.csv("user_behavior_dataset.csv")
head(df)
```

```{r}
df <- df[, !names(df) %in% "User.ID"]
tail(df)
```

```{r}
dim(df)
```

```{r}
nrow(df)
ncol(df)
```

```{r}
str(df)
```

```{r}
summary(df)
```

```{r}
describe(df)

```

```{r}
colSums(is.na(df))
```

```{r}
sum(duplicated(df))

```

## Explanatory Data Analysis

```{r}
# Distribution of User Behavior Classes
ggplot(df, aes(x = as.factor(User.Behavior.Class))) +
  geom_bar(fill = "skyblue") +
  labs(title = "Distribution of User Behavior Classes",
       x = "User Behavior Class",
       y = "Count") +
  theme_minimal()
```

The graph shows the distribution of User Behavior Classes, ranging from 1 (light usage) to 5 (extreme usage). The counts across the classes are fairly balanced, with Class 2 (light-to-moderate usage) having the highest number of users. Classes 3, 4, and 5, representing increasing levels of usage intensity, have slightly fewer users but remain consistent in size. Interestingly, Class 1 (light usage) also holds a significant proportion, suggesting that a notable portion of users exhibit minimal device usage. Overall, the balanced distribution across all classes ensures reliable insights for understanding and predicting user behavior trends.


```{r}
# The distribtuion of age by User Behavior Class
age_behavior_plot <- ggplot(df, aes(x = factor(User.Behavior.Class), y = Age)) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Age Distribution by User Behavior Class", x = "User Behavior Class", y = "Age") +
  theme_minimal()
age_behavior_plot
```

This boxplot shows the age distribution for each User Behavior Class from light (1) to extreme (5) usage. The median age for all classes is around 40 years, with no big differences between the groups. Class 1 (light usage) has a slightly younger median and less variation compared to the other classes. Overall, the ages in all classes range from about 20 to 60 years, showing that user behavior doesn't change much with age.

```{r}
# The distribution of gender analysis
ggplot(df, aes(x = Gender, fill = Gender)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set2") +  # Use a different palette (e.g., Set2)
  labs(title = "Gender Analysis", 
       x = "Gender", 
       y = "Total Devices") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5)) +
  theme(plot.title = element_text(size = 16, hjust = 0.5))

```

The bar chart shows the number of users by gender. There are slightly more male users than female users in this dataset. Both groups have similar totals, with males having a small lead. This means the data is fairly even, which is good for comparing behavior between genders.

```{r}
# The distribution of operating sytem analysis
ggplot(df, aes(x = Operating.System, fill = Operating.System)) +
  geom_bar() +
  scale_fill_viridis(discrete = TRUE) +  # Viridis for discrete categories
  labs(title = "Operating System Analysis", 
       x = "Operating System", 
       y = "Total Devices") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(plot.title = element_text(size = 16, hjust = 0.5))

```

The bar chart shows the number of users for each Operating System. Most users in this dataset use Android, while fewer users use iOS. The difference is large, with Android having more than twice as many users as iOS. This suggests that Android devices are much more common among the users in this dataset.

```{r}
# Correlation heatmap
df$Gender <- as.factor(df$Gender)
cor_data <- df %>%
  select(App.Usage.Time..min.day., Screen.On.Time..hours.day., Battery.Drain..mAh.day., 
         Number.of.Apps.Installed, Data.Usage..MB.day., Age)
cor_matrix <- cor(cor_data)
corr_plot <- ggcorrplot(cor_matrix, lab = TRUE, title = "Correlation Heatmap")
corr_plot
```

The correlation heatmap shows how different variables are related to each other. App Usage Time, Screen-On Time, Battery Drain, Number of Apps Installed, and Data Usage all have very strong positive correlations (close to 1), meaning they increase together. However, Age has no clear relationship with any of the other variables, as its correlation values are close to zero. Overall, the strong relationships among App Usage Time (minutes per day), Screen-On Time (hours per day), Battery Drain (mAh per day), Number of Apps Installed, and Data Usage (MB per day) suggest that these metrics are closely connected to how users interact with their devices.

```{r}
# Screen-On Time vs User Behavior Class
ggplot(df, aes(x = Screen.On.Time..hours.day., y = as.factor(User.Behavior.Class))) +
  geom_boxplot(fill = "lightgreen") +
  labs(title = "Screen-On Time vs User Behavior Class",
       x = "Screen-On Time (hours/day)",
       y = "User Behavior Class") +
  theme_minimal()

```

The boxplot shows the relationship between Screen-On Time and User Behavior Class. As the User Behavior Class increases from 1 (light usage) to 5 (extreme usage), the Screen-On Time also increases. Users in Class 1 have the lowest Screen-On Time, around 1 to 2 hours per day, while users in Class 5 have the highest Screen-On Time, around 10 to 12 hours per day. It suggests that Screen-On Time is a strong indicator of user behavior, with higher usage classes spending more time on their devices.

```{r}
# Battery Drain vs User Behavior Class
ggplot(df, aes(x = Battery.Drain..mAh.day., y = as.factor(User.Behavior.Class))) +
  geom_boxplot(fill = "lightcoral") +
  labs(title = "Battery Drain vs User Behavior Class",
       x = "Battery Drain (mAh/day)",
       y = "User Behavior Class") +
  theme_minimal()
```

The boxplot shows the relationship between Battery Drain and User Behavior Class. As the User Behavior Class increases from 1 (light usage) to 5 (extreme usage), the Battery Drain also increases. Users in Class 1 have the lowest Battery Drain, around 300 to 600 mAh/day, while users in Class 5 have the highest Battery Drain, around 2500 to 3000 mAh/day. It suggests that higher usage behavior leads to more battery consumption, which makes sense as increased activity drains the battery faster.

```{r}
# Data Usage vs User Behavior Class
ggplot(df, aes(x = Data.Usage..MB.day., y = as.factor(User.Behavior.Class))) +
  geom_boxplot(fill = "gold") +
  labs(title = "Data Usage vs User Behavior Class",
       x = "Data Usage (MB/day)",
       y = "User Behavior Class") +
  theme_minimal()

```

The boxplot shows the relationship between Data Usage and User Behavior Class. As the User Behavior Class increases from 1 (light usage) to 5 (extreme usage), the Data Usage also increases. Users in Class 1 use the least data, around 100 to 300 MB/day, while users in Class 5 have the highest data usage, around 2000 to 2500 MB/day. It suggests that users with higher behavior classes consume more data daily, likely due to increased app usage, streaming, or other online activities.

```{r}
# Number of Apps Installed vs User Behavior Class
ggplot(df, aes(x = Number.of.Apps.Installed, y = as.factor(User.Behavior.Class))) +
  geom_boxplot(fill = "purple") +
  labs(title = "Number of Apps Installed vs User Behavior Class",
       x = "Number of Apps Installed",
       y = "User Behavior Class") +
  theme_minimal()

```

The boxplot shows the relationship between the Number of Apps Installed and User Behavior Class. As the User Behavior Class increases from 1 (light usage) to 5 (extreme usage), the number of apps installed also increases. Users in Class 1 have the fewest apps, around 10 to 20, while users in Class 5 have the most, around 80 to 100 apps. It suggests that users with higher behavior classes tend to install and use more apps, which is likely tied to their heavier device usage.

```{r}
columns <- c('App.Usage.Time..min.day.', 'Screen.On.Time..hours.day.',
             'Battery.Drain..mAh.day.', 'Number.of.Apps.Installed',
             'Data.Usage..MB.day.', 'Age')

# Create individual histograms for each column
plots <- lapply(columns, function(col) {
  ggplot(df, aes_string(x = col)) +
    geom_histogram(aes(y = ..density..), bins = 30, fill = "green", color = "black", alpha = 0.7) +
    geom_density(color = "darkgreen") +
    labs(title = col, x = gsub("\\.\\.", " ", col), y = "Density") +
    theme_minimal()
})

combined_plot <- Reduce(`+`, plots) + plot_layout(ncol = 3)
combined_plot
```

The histograms show the distribution of six variables: App Usage Time, Screen-On Time, Battery Drain, Number of Apps Installed, Data Usage, and Age. Most variables, like App Usage Time, Screen-On Time, and Data Usage, are skewed to the right, meaning many users have lower values while a few have very high values. Battery Drain and Number of Apps Installed are more spread out across their ranges, with no clear concentration in a specific area. The Age distribution is fairly uniform, showing a balanced spread of users across different age groups. It suggests that most usage metrics vary widely, with certain users driving the higher values.

```{r}
# Distribution of User Behavior Class by Gender
ggplot(df, aes(x = as.factor(User.Behavior.Class), fill = Gender)) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of User Behavior Classes by Gender",
       x = "User Behavior Class",
       y = "Count",
       fill = "Gender") +
  theme_minimal()
```

The bar chart shows the distribution of User Behavior Classes by gender, where classes range from 1 (light usage) to 5 (extreme usage). Males generally have higher counts in most classes compared to females. In Class 1 (light usage), males slightly outnumber females. Class 3 (moderate usage) shows a balanced count between males and females. However, in Class 5 (extreme usage), males have a noticeably higher count, suggesting that males are more likely to fall into the heavy usage categories.

```{r}
# Screen-On Time vs User Behavior Class by Gender
ggplot(df, aes(x = as.factor(User.Behavior.Class), y = Screen.On.Time..hours.day., fill = Gender)) +
  geom_boxplot() +
  labs(title = "Screen-On Time by User Behavior Class and Gender",
       x = "User Behavior Class",
       y = "Screen-On Time (hours/day)",
       fill = "Gender") +
  theme_minimal()
```

The boxplot shows the Screen-On Time by User Behavior Class and Gender. As the User Behavior Class increases from 1 (light usage) to 5 (extreme usage), Screen-On Time also increases for both genders. In Class 1 and 2, males and females have similar Screen-On Time, ranging between 2 to 4 hours/day. However, in Class 5 (extreme usage), females tend to have slightly higher Screen-On Time, with values reaching around 12 hours/day, compared to males. This suggests that in heavier usage classes, Screen-On Time increases significantly, with slight gender differences.

```{r}

# Screen-On Time vs User Behavior Class by Age Group
df$Age.Group <- cut(df$Age, 
                    breaks = c(18, 25, 35, 45, 60), 
                    labels = c("18-25", "26-35", "36-45", "46-59"),
                    include.lowest = TRUE)

ggplot(df, aes(x = as.factor(User.Behavior.Class), 
               y = Screen.On.Time..hours.day., 
               fill = Age.Group)) +
  geom_boxplot() +
  labs(title = "Screen-On Time by User Behavior Class and Age Group",
       x = "User Behavior Class",
       y = "Screen-On Time (hours/day)",
       fill = "Age Group") +
  theme_minimal()
```

## Conclusion

The exploratory data analysis (EDA) of the dataset reveals significant trends in mobile device usage behavior across five user behavior classes ranging from light to extreme usage. Key findings show that as the behavior class increases, Screen-On Time, App Usage Time, Battery Drain, Number of Apps Installed, and Data Usage also increase, suggesting a strong connection between user behavior intensity and these metrics. Gender differences were observed, with males generally dominating higher behavior classes, while females in Class 5 displayed slightly higher Screen-On Time. Age, however, showed no significant relationship with any usage metrics.

For further investigation, a predictive model could be developed to classify users into behavior classes based on device usage metrics. Techniques such as logistic regression, decision trees, or machine learning models (e.g., Random Forest, XGBoost) can be applied to identify key predictors of behavior classes. Additionally, further analysis can explore time-based trends (e.g., daily or hourly usage patterns) and investigate how specific app categories contribute to behavior intensity. This work provides a strong foundation for understanding user behavior and optimizing strategies for app developers, device manufacturers, or digital well-being initiatives.
